---
title: "Homework Feb. 5"
output: pdf_document
---

## Univariate Assignment

Read in tree data, metadata can be found in 
`./data/tree_metadata.txt`


1. Carry out an exploratory analysis using the tree dataset. For each species  
address the following additional questions:

  * how well does the exploratory model appear to explain cover?
    * which explanatory variables are the most important?
    * do model diagnostics indicate any problems with violations of
      OLS assumptions?
    * are you able to explain variance in one species better than another? 
    
```{r}
library(car)
trees = read.csv('http://dmcglinn.github.io/quant_methods/data/treedata_subset.csv')
head(trees)
levels(trees$species)

#Subset Rubrum

disturb<-as.factor(trees$disturb)
rubrum<-trees[trees$species=="Acer rubrum",]
rubrum
R<-lm(cover~elev+disturb+tci+streamdist+beers, data=rubrum)


summary(R)
coefficients(R)
plot(R)


#aov

Raov<- Anova(R, type=3)
Raov
summary(R)
par(mfrow=c(2,2))
plot(R)


#Look at Rubrum Residuals


Rubresid<-residuals(R)
qqnorm(Rubresid)
qqline(Rubresid)


#Assessment of the Linear Model Assumptions
library(gvlma)
gvmodelR<- gvlma(R) 
gvmodelR


```
  
    
Rubrum Analysis: 
For the rubrum species, we can see that the model does an okay job to 
explain cover, but we can do better- I say this because when we look at the 
residuals (scale-location of fitted values to sqrt of standardized residuals), 
we can see trends in our residuals, and in that case, our model has room for 
improvement.. We can also look at our adjusted R squared value and it is
extremely small, which can be improved with a better model. It currently explains
only 3% of the variance.
We can see that elevation, streamdist, beers, and the CORPLOG disturbance 
(intercept) are important in determining cover. This model does a 
pretty good job of explaining cover, because when we look at a qqplot of the 
residuals, we can really see that there is not many outliars (and the outliars
that we do have don't have much leverage at all-hardly any, actually..)
The data are normally distributed and have equal levels of variance. 

Out of all of the variables that account for rubrum cover area, it appears that 
disturbance (excluding CORPLOG) and tci are the only factors that do not 
affect the cover of Acer rubrum. 
This makes sense, because if we know that a priori 
that this tree is a generalist, it may be able to withstand a variety of
disturbance types (so that disturbance has no effect). 
So I feel that this does a pretty decent job at explaining cover, with the most 
important variables being elevation, streamdist, beers. 

Using the gvlma command, we can see what assumptions the model violates. For 
rubrum, it violates skewness, kurtosis, and global stat. So the data are 
extremely peaked comapred to a normal dist. and they is asymmetry in the 
probability distribution. 


```{r}
#Subset Frasier

fraseri<-trees[trees$species=="Abies fraseri",]
fraseri
F<-lm(cover~elev+disturb+tci+streamdist+beers, data=fraseri)


summary(F)
coefficients(F)
plot(F)



#aov

disturb<-as.factor(trees$disturb)

Faov<-Anova(F, type=3)
Faov
par(mfrow=c(2,2))
plot(F)


#Look at Frasier Residuals


frasresid<-residuals(F)
qqnorm(frasresid)
qqline(frasresid)


#Model Violations
gvmodelF<- gvlma(F) 
gvmodelF


```


Frasier Analysis: The variables that are the most important for frasier cover
are the CORPLOG (intercept), and elevation (both of these also had an effect
on the rubrum cover as well.), however the other variables had no significant
effect on cover. 
This model does a better job of fitting to the data than it did for rubrum,
and we can see that this model violates fewer assumptions that it did for rubrum
. Here, we can see that the only violation that this model does not follow
is that for heteroscedasticity. Looking at the adjusted R squared, we can 
see that this model also does a much better fitting the frasier data than it 
did for the rubrum data, because now it explains about 50% of the data. 
So this model specifically does a much better job of explaining the frasier 
data set than the rubrum data set. 
These data do not appear to be normally distributed, and there is more 
variance is occurring here (there are more outliers in the qq plot.). However, 
there are much fewer data points so these outliars have a bit more leverage
on the qqline. When we look at the residuals vs. fitted, we can see
heteroscedasticity. 



#-----------------------------------------------------------------------------
2. You may have noticed that the variable cover is defined as 
positive integers between 1 and 10. and is therefore better treated
as a discrete rather than continuous variable. 
Re-examine your solutions to the question above but from the
perspective of a General Linear Model (GLM) with a Poisson error term
(rather than a Gaussian one as in OLS). 
The Poisson distribution generates integers 0 to positive infinity
so this may provide a good first approximation. 
Your new model calls will look as follows:


    ```{r}
disturb<-as.factor(trees$disturb)

R
F

summary(R)
coefficients(R)
plot(R)

summary(F)
coefficients(F)
plot(F)


acer_glm<-glm(cover~elev+disturb+tci+streamdist+beers, data=rubrum, family='poisson')
fras_glm<-glm(cover~elev+disturb+tci+streamdist+beers, data=fraseri, family='poisson')

    
summary(acer_glm)    
summary(fras_glm)

    ```

We can see that for rubrum, the same factors (CORPLOG, elev, streamdist, and 
beers) for both models shows these affect the cover. This follows for the same
as frasier.  We are able to explain variance much better in our frasier 
species at 60% and only 4% of the variance in rubrum. These models are better 
fits for both data sets because the glm models explain more variance than 
linear models. 




For assessing the degree of variation explained you can use a 
pseudo-R-squared statistic (note this is just one of many possible)

```{r}

pseudo_r2 = function(glm_mod) {
    1 -  glm_mod$deviance / glm_mod$null.deviance
}

pseudo_r2(acer_glm)
#0.03997917
pseudo_r2(fras_glm)
#0.60931
```

0.6091592 (61%) variance explained for the frasier general linear model 
compared to the 50% variance explained for the OLS model. This model improves
the variance explained by 10% which rocks.


However itonly explains 4% of the variance for the rubrum glm, which is hardly
a difference from the multiple r^2 value we found in the OLS model of about
3.5%..so it imporved it slightly but not by much to really  note.
    
   


Compare the residual sums of squares between the traditional OLS 
and glm models using `anova` (Note: not `Anova`) as such

    ```{r}
    anova(R, acer_glm)
    
    
    anova(F, fras_glm)
    
    ```

A small residual sums of squares indicates which model has a tighter/better
fit to the data set. In both cases for both trees, the general linear model 
fits much better, because these models have a significantly lower RSS value.




Does it appear that changing the error distribution changed the
results much? In what ways? 

This accounts for the samples sizes in each of the groups and now in the 
glm models the standard errors are now much smaller, which is more representativ
e of the model overall,which results in a more reliable p-value we can trust. 
Looking at the acer rubrum, in the linear model we can see that the tci
shows an alpha at .07 which gives it some noise on how much variance it explains
but when we adjust for the sample sizes , the glm model shows that the 
alpha changes to .112.  



3. Provide a plain English summary (i.e., no statistics) of what you have
found and what conclusions we can take away from your analysis?


We looked at 2 species of trees, rubrum and frasier, one being a specialist
and the other being a generalist. We would expect different variables to 
affect the distribution of these trees, so we wanted to look at different models
to see which could explain the most variation. The easiest would be to look 
at a linear model with multiple independent variables. This model did a pretty
good job at explaining frasier, but not a good job for rubrum. Models that 
do a good job explaining a majority of the variation can be considered a good
model to follow. So we then wanted to look to see if there was a better model to
follow. So we looked at a general linear model which allows for the 
response variable to have an error distribution other than a normal distribution
. The GLM generalizes linear regression by allowing the linear model to be 
related to the response variable, which we can see by the amount of variation
that each independent variable contributes. 
From this analysis, we can see that the models are going to do a better job
of explaining variation for the frasier trees, because these are a specialist 
tree. Meaning that there is going to be a few factors that will prohibit its
distribution, which we can see is elevation and a type of disturbance. We 
see that either model, on the flip side, still does not do a good job of 
explaining the variation of the data. However, this may arrise from the 
problem of the data set. A priori we know that the rubrum tree is a generalist,
so it is going to take a greater range of predictor variables to affect 
cover. It appears that if more/ different predictor variables where used in the
model, the model could have done a better job of explaining the cover of the 
acer rubrum species. With that being said, the predictor variables that do 
limit the cover of the rubrum tree might have not been included. But 
we only know this post-analysis, so I would still trust the model because it 
is able to explain 60% of the variation for the frasier fur. We can take away
that for both trees, elevation and CORGPLOG have a good influence on the 
cover. 
To visualize the amount each variable influences cover, I would have probably
 done a Tukey test, just to see the variation and influence (visually) each 
variable had on cover.
The poisson distribution is more applicable because the response variable 
is considered to be categorical data, rather than continuous data. It is used
to model the number of events occuring within a certain time, and 
different predictor variables. Since our response is not continuous (which
a lm would be better to use if so), the poisson dist is better to use.
